Extended Data

### 2021-10-15 doc/manuscript/Dow_manuscript_Nature.Rmd
discussion

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance. 
Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`), is likely driven by cues other than carbon limitation, such as daylength<!--REF?--> or sink limitation[@zani_increased_2020].
Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.

<!-- it remains unclear where extra carbon goes (1 par, maybe 2)-->
Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig. `r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that this extra C sequestration in years with warm springs does not substantially augment woody growth, raising the question of where it goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Stem growth is more sensitive to the water deficits that may be associated with high temperatures than is photosynthesis[@muller_water_2011], and hot or dry conditions have been observed to shift allocation towards roots[@zohner_how_2021], reproduction[@hacket-pain_drought_2017], or respiration[**REFS**].
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], consistent with studies focused on higher-latitude forests that show enhanced C sequestration in years with warm springs to be offset by higher C losses in the autumn[@piao_net_2008; @barichivich_large-scale_2013].
Notably, though, turnover of extra C fixed during years with warm springs may occur with sufficient temporal lag that it would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth probably has a residence time of more than a year within the ecosystem; however, the amount and fate of this carbon remains unknown.

>>>>

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance. 
Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength<!--REF?--> or sink limitation[@zani_increased_2020].
Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with of drought stress[@dorangeville_drought_2018].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.

<!-- it remains unclear where extra carbon goes (1 par, maybe 2)-->
Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig. `r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
We show that this extra C sequestration in years with warm springs does not substantially augment woody growth, raising the question of where it goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

A second possibility is that any additional C fixed during years with warm springs may be allocated, potentially in the following growing season, to plant functions other than stem growth, including respiration, reproduction, and production of foliage, fine roots[@zohner_how_2021], or root exudates. 
Most of this carbon would have a shorter residence time within the ecosystem than C allocated to woody productivity[**REF**], <!-- [@trumbore_quantifying_2002].?-->but would likely remain in the ecosystem for more than a year<!--ref-->, such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year growth has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.
